Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-03-27 document/ms.tex
typo
order of period and radius

\begin{abstract}
Although no true extra-solar Earth analog is known, hundreds of planets have
been found around Sun-like stars that are either Earth-sized but on shorter
periods, or else on year-long orbits but somewhat bigger than Earth.
These populations ought to permit a (possibly highly uncertain) extrapolated
measurement of the probability or rate at which Sun-like stars host true Earth
analogs.
Here we build a set of probabilistic models of the population of known small
planets around G dwarfs.
The models all take into account the transit probability and the completeness
(or detection efficiency) as a function of planet radius and orbital period.
Because we forward model the data, we do not ever have to perform the
traditional weighting of the data by the inverse detection efficiency.
Our most sophisticated models do not assume that the joint radius--period
distribution is separable, and correctly take into account the significant
observational uncertainties on the planet radii, using the machinery of
hierarchical inference.
We reconsider and confirm with weaker assumptions prior results on the
distribution of exoplanet radii that show a flattening of the distribution
below a few Earth radii.
One output of our models is fully marginalized estimates---marginalizing out
observational uncertainties and all distribution parameters---of the rate
density of true Earth analogs, expressed as a number of planets per star per
natural logarithmic interval of radius and period, evaluated at the properties
of Earth.
The rate density we infer is an extrapolation; it depends strongly on the
permitted flexibility of the distribution model, but it tends to come out
at around ten percent, with large uncertainty.
\end{abstract}

@@

This study is novel in a number of ways.
We create probabilistic information about the full exoplanet population
\emph{not} by weighting the detected objects by the inverse detection
efficiency (\citealt{dong, petigura}), but rather by forward modeling the
observed distribution subject to the detection efficiency.
\hoggtodo{Check that Dong \& Zhu do indeed weight by inverse probability.}
Re-weighting the data by inverse selection probability---a method called
V-max in the quasar and galaxy luminosity function contexts---is not wrong,
but it produces a higher-variance estimate of the population than a justified
likelihood approach.
One simple intuition along these lines is that the inverse probability weights
can be large in dynamic range, so the effective number of samples, after
weighting, can be much smaller than the actual number of samples.
Another is that it tends to upweight most strongly the most marginal
detections in the data sample, since these tend to be the samples that lie in
the region of lowest completeness.  In this work, we build a justified
likelihood of the observed data, taking into account the completeness function
or detection efficiency.
Our likelihood function is a variable-rate Poisson likelihood; it makes the
fundamental assumption that the data points are independently drawn from the
model, but is agnostic about all other properties of the data or distribution
functions.
Indeed, the assumptions behind this study are weaker than those of any
previous study of the radius and period distribution of the Earth-analog
population (\citealt{dong, petigura}).

Another new aspect of this study is that it makes use of very flexible
distribution function models.
Indeed most of the models we consider are ``non-parametric'' in the weak sense
that they have enormous numbers of free parameters (not the strong sense of
having an \emph{infinite} number of free parameters).
The models are protected from degeneracies and over-fitting by priors.
In some models we assume that the two-dimensional distribution of planets in
radius and period is separable; in others we don't.
We treat all of these parameters as ``nuisance parameters'' when we ask
questions about the occurrence of Earth analogs.
That is, our rate conclusions are fully marginalized, and those
marginalizations are over non-trivial numbers of parameters.

@@

From our perspective, a ``model'' is a likelihood function---a probability
density (or distribution) function (PDF) for the data given
parameteters---and a set of prior PDFs for the parameters.
In what follows, we will treat as ``data'' a certain set of exoplanet radius
and period measurements (\citealt{petigura}).
We will treat the uncertainties on those measurements as ``prior
information''; that is, we will treat those as true as given and not as being
generated by the parameterized likelihood function.

In the models that follow, the most important parameters will be parameters
of the PDF for the exoplanet radii and periods.
However, there will also be ``latent parameters'' that represent the ``true''
values of the exoplanet radii and periods.
That is, for each exoplanet, there is a ``measured'' or ``observed'' radius
and period, which constitute our data.
These relate to the unobserved ``true'' radius and period by a noise model.
For the conclusions of this study, the true values of radius and period are
nuisance parameters; we will marginalize them out.

@@

We have made the first measurement of the Earth-like exoplanet population that
is based on forward modeling, takes proper account of the observational
uncertainties on planet radii, and makes use of extremely flexible
non-separable distributions for radius and period.
We find \hoggtodo{BLAH BLAH BLAH.}

Although our method makes weaker assumptions than any previous work in this
area, of course it \emph{does} make strong assumptions.
It assumes that the data set (from \citealt{petigura}) contains planets and
only planets (no false positives).
It assumes that the individual-planet radius uncertainties are also correctly
determined.
It assumes that the completeness calculation (again from \citealt{petigura})
is accurate and smoothly varying with radius and period; it makes no attempt
to adjust the completeness star-by-star for brightness or variability.
It assumes that the exoplanets are all independently drawn from the same
distribution function, with no regard to age, metallicity, or multiplicity.

Although we have taken account of the radius uncertainties in a principled
way, we did make one important approximation: We ignored planets with central
values far outside the radius and period box in which we built the model.
That is, we assumed that there are no very large outliers in radius
determination.
This assumption is a complex one; it involves the effective, end-to-end error
distribution in the data set, but also smoothness of the distribution function
at or near the edges of the analysis box.

@@

\newcommand{\arxiv}[1]{\href{http://arxiv.org/abs/#1}{arXiv:#1}}
\begin{thebibliography}{}\raggedright
\bibitem[Brewer \& Stello(2009)]{brewer}
Brewer, B.~J., \& Stello, D.\ 2009, \mnras, 395, 2226 (\arxiv{0902.3907})
\bibitem[Carter \& Winn(2009)]{carter}
Carter, J.~A., \& Winn, J.~N.\ 2009, \apj, 704, 51 (\arxiv{0909.0747})
\bibitem[Dong \& Zhu(2013)]{dong}
Dong, S., \& Zhu, Z.\ 2013, \apj, 778, 53 (\arxiv{1212.4853})
\bibitem[Hogg et \etal(2013)]{hoggwhitepaper}
Hogg, D.~W., Angus, R., Barclay, T., et al.\ 2013, \arxiv{1309.0653}
\bibitem[Hogg \etal(2010)]{hogge}
Hogg, D.~W., Myers, A.~D., \& Bovy, J.\ 2010, \apj, 725, 2166 (\arxiv{1008.4146})
\bibitem[Petigura \etal(2013)]{petigura}
Petigura, E.~A., Howard, A.~W., \& Marcy, G.~W.\ 2013,
Proceedings of the National Academy of Science, 110, 19273 (\arxiv{1311.6806})
\bibitem[Roberts \etal(2013)]{roberts}
Roberts, S., McQuillan, A., Reece, S., \& Aigrain, S.\ 2013, \mnras, 435, 3639 (\arxiv{1308.3644})
\end{thebibliography}

>>>>

\begin{abstract}
Although no true extra-solar Earth analog is known, hundreds of planets have
been found around Sun-like stars that are either Earth-sized but on shorter
periods, or else on year-long orbits but somewhat bigger than Earth.
These populations ought to permit a (possibly highly uncertain) extrapolated
measurement of the probability or rate at which Sun-like stars host true Earth
analogs.
Here we build a set of probabilistic models of the population of known small
planets around G dwarfs.
The models all take into account the transit probability and the completeness
(or detection efficiency) as a function of planet radius and orbital period.
Because we forward model the data, we do not ever have to perform the
traditional weighting of the data by the inverse detection efficiency.
Our most sophisticated models do not assume that the joint radius--period
distribution is separable, and correctly take into account the significant
observational uncertainties on the planet radii, using the machinery of
hierarchical inference.
We reconsider and confirm with weaker assumptions prior results on the
distribution of exoplanet radii that show a flattening of the distribution
below a few Earth radii.
One output of our models is fully marginalized estimates---marginalizing out
observational uncertainties and all distribution parameters---of the rate
density of true Earth analogs, expressed as a number of planets per star per
natural logarithmic interval of period and radius, evaluated at the properties
of Earth.
The rate density we infer is an extrapolation; it depends strongly on the
permitted flexibility of the distribution model, but it tends to come out
at around ten percent, with large uncertainty.
\end{abstract}

@@

This study is novel in a number of ways.
We create probabilistic information about the full exoplanet population
\emph{not} by weighting the detected objects by the inverse detection
efficiency (\citealt{dong, petigura}), but rather by forward modeling the
observed distribution subject to the detection efficiency.
\hoggtodo{Check that Dong \& Zhu do indeed weight by inverse probability.}
Re-weighting the data by inverse selection probability---a method called
V-max in the quasar and galaxy luminosity function contexts---is not wrong,
but it produces a higher-variance estimate of the population than a justified
likelihood approach.
One simple intuition along these lines is that the inverse probability weights
can be large in dynamic range, so the effective number of samples, after
weighting, can be much smaller than the actual number of samples.
Another is that it tends to upweight most strongly the most marginal
detections in the data sample, since these tend to be the samples that lie in
the region of lowest completeness.  In this work, we build a justified
likelihood of the observed data, taking into account the completeness function
or detection efficiency.
Our likelihood function is a variable-rate Poisson likelihood; it makes the
fundamental assumption that the data points are independently drawn from the
model, but is agnostic about all other properties of the data or distribution
functions.
Indeed, the assumptions behind this study are weaker than those of any
previous study of the period and radius distribution of the Earth-analog
population (\citealt{dong, petigura}).

Another new aspect of this study is that it makes use of very flexible
distribution function models.
Indeed most of the models we consider are ``non-parametric'' in the weak sense
that they have enormous numbers of free parameters (not the strong sense of
having an \emph{infinite} number of free parameters).
The models are protected from degeneracies and over-fitting by priors.
In some models we assume that the two-dimensional distribution of planets in
period and radius is separable; in others we don't.
We treat all of these parameters as ``nuisance parameters'' when we ask
questions about the occurrence of Earth analogs.
That is, our rate conclusions are fully marginalized, and those
marginalizations are over non-trivial numbers of parameters.

@@

From our perspective, a ``model'' is a likelihood function---a probability
density (or distribution) function (PDF) for the data given
parameteters---and a set of prior PDFs for the parameters.
In what follows, we will treat as ``data'' a certain set of exoplanet period
and radius measurements (\citealt{petigura}).
We will treat the uncertainties on those measurements as ``prior
information''; that is, we will treat those as true as given and not as being
generated by the parameterized likelihood function.

In the models that follow, the most important parameters will be parameters
of the PDF for the exoplanet periods and radii.
However, there will also be ``latent parameters'' that represent the ``true''
values of the exoplanet periods and radii.
That is, for each exoplanet, there is a ``measured'' or ``observed'' period
and radius, which constitute our data.
These relate to the unobserved ``true'' period and radius by a noise model.
For the conclusions of this study, the true values of period and radius are
nuisance parameters; we will marginalize them out.

@@

We have made the first measurement of the Earth-like exoplanet population that
is based on forward modeling, takes proper account of the observational
uncertainties on planet radii, and makes use of extremely flexible
non-separable distributions for period and radius.
We find \hoggtodo{BLAH BLAH BLAH.}

Although our method makes weaker assumptions than any previous work in this
area, of course it \emph{does} make strong assumptions.
It assumes that the data set (from \citealt{petigura}) contains planets and
only planets (no false positives).
It assumes that the individual-planet radius uncertainties are also correctly
determined.
It assumes that the completeness calculation (again from \citealt{petigura})
is accurate and smoothly varying with period and radius; it makes no attempt
to adjust the completeness star-by-star for brightness or variability.
It assumes that the exoplanets are all independently drawn from the same
distribution function, with no regard to age, metallicity, or multiplicity.

Although we have taken account of the radius uncertainties in a principled
way, we did make one important approximation: We ignored planets with central
values far outside the period and radius box in which we built the model.
That is, we assumed that there are no very large outliers in radius
determination.
This assumption is a complex one; it involves the effective, end-to-end error
distribution in the data set, but also smoothness of the distribution function
at or near the edges of the analysis box.

@@

\newcommand{\arxiv}[1]{\href{http://arxiv.org/abs/#1}{arXiv:#1}}
\begin{thebibliography}{}\raggedright
\bibitem[Brewer \& Stello(2009)]{brewer}
Brewer, B.~J., \& Stello, D.\ 2009, \mnras, 395, 2226 (\arxiv{0902.3907})
\bibitem[Carter \& Winn(2009)]{carter}
Carter, J.~A., \& Winn, J.~N.\ 2009, \apj, 704, 51 (\arxiv{0909.0747})
\bibitem[Dong \& Zhu(2013)]{dong}
Dong, S., \& Zhu, Z.\ 2013, \apj, 778, 53 (\arxiv{1212.4853})
\bibitem[Hogg \etal(2013)]{hoggwhitepaper}
Hogg, D.~W., Angus, R., Barclay, T., et al.\ 2013, \arxiv{1309.0653}
\bibitem[Hogg \etal(2010)]{hogge}
Hogg, D.~W., Myers, A.~D., \& Bovy, J.\ 2010, \apj, 725, 2166 (\arxiv{1008.4146})
\bibitem[Petigura \etal(2013)]{petigura}
Petigura, E.~A., Howard, A.~W., \& Marcy, G.~W.\ 2013,
Proceedings of the National Academy of Science, 110, 19273 (\arxiv{1311.6806})
\bibitem[Roberts \etal(2013)]{roberts}
Roberts, S., McQuillan, A., Reece, S., \& Aigrain, S.\ 2013, \mnras, 435, 3639 (\arxiv{1308.3644})
\end{thebibliography}
